---
title: "wolf_analysis"
output: pdf_document
---

##Load libraries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readxl)
library(dplyr)
library(ggplot2)

```

##Load datasets
```{r cars}

animal_id_sheet = read_xlsx("/Users/MichaelGrodecki/Downloads/CS Stuff/LISA/Resources/wolf_analysis/Wolf_data-LISA.xlsx", sheet = "animal id")
ethogram_sheet = read_xlsx("/Users/MichaelGrodecki/Downloads/CS Stuff/LISA/Resources/wolf_analysis/Wolf_data-LISA.xlsx", sheet = "Ethogram")
ethogram_revised_sheet = read_xlsx("/Users/MichaelGrodecki/Downloads/CS Stuff/LISA/Resources/wolf_analysis/Wolf_data-LISA.xlsx", sheet = "Ethogram Revised")
activity_budget_sheet = read_xlsx("/Users/MichaelGrodecki/Downloads/CS Stuff/LISA/Resources/wolf_analysis/Wolf_data-LISA.xlsx", sheet = "activity budget")
all_occurance_sheet = read_xlsx("/Users/MichaelGrodecki/Downloads/CS Stuff/LISA/Resources/wolf_analysis/Wolf_data-LISA.xlsx", sheet = "all occurance")

animal_id_sheet
ethogram_sheet
ethogram_revised_sheet
activity_budget_sheet
all_occurance_sheet

```

## Data Visualization of bold and timid behaviors - bar chart, pie chart, scatterplot 
```{r}

# List of bold behaviors
bold_behaviors <- c("Bold approach- fenceline", 
                    "Bold approach- conspecific", 
                    "Bold approach- guest")

#List of timid behaviors
timid_behaviors <- c("Timid approach- fenceline",
                    "Fearful flight/startle",
                    "Flight or startle")

# Count observations of bold behaviors
bold_count <- activity_budget_sheet %>%
  filter(Behavior %in% bold_behaviors) %>%
  summarise(count = n()) %>%
  pull(count)

# Count observations of timid behaviors
timid_count <- activity_budget_sheet %>%
  filter(Behavior %in% timid_behaviors) %>%
  summarise(count = n()) %>%
  pull(count)

# Display the counts
cat("Bold Behavior Observations: ", bold_count, "\n")
####Output: Bold Behavior Observations:  59 

cat("Timid Behavior Observations: ", timid_count, "\n")
####Output: Timid Behavior Observations:  13 

# Filter rows with bold behaviors
bold_data <- activity_budget_sheet %>%
  filter(Behavior %in% bold_behaviors)

# Filter rows with timid behaviors
timid_data <- activity_budget_sheet %>%
  filter(Behavior %in% timid_behaviors)

# Create a summary of counts for bold behaviors by color
bold_summary <- bold_data %>%
  group_by(color, Behavior) %>%
  summarise(count = n(), .groups = "drop")

# Create a summary of counts for timid behaviors by color
timid_summary <- timid_data %>%
  group_by(color, Behavior) %>%
  summarise(count = n(), .groups = "drop")

# Visualize the Bold relationship with coat color - bar chart
ggplot(bold_summary, aes(x = color, y = count, fill = Behavior)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Relationship Between Coat Color and Bold Behaviors",
       x = "Color",
       y = "Coat Count of Bold Behaviors",
       fill = "Bold Behavior") +
  theme_minimal()

# Visualize the Timid relationship with coat color - bar chart
ggplot(timid_summary, aes(x = color, y = count, fill = Behavior)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Relationship Between Coat Color and Timid Behaviors",
       x = "Coat Color",
       y = "Count of Timid Behaviors",
       fill = "Timid Behavior") +
  theme_minimal()

# Summarize total counts of bold behaviors by Color - pie chart
bold_summary_pie <- bold_data %>%
  group_by(color) %>%
  summarise(total_count = n(), .groups = "drop")

# Summarize total counts of timid behaviors by Color - pie chart
timid_summary_pie <- timid_data %>%
  group_by(color) %>%
  summarise(total_count = n(), .groups = "drop")

# Visualize the relationship between coat color and boldness - Pie chart
ggplot(bold_summary_pie, aes(x = "", y = total_count, fill = color)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Proportion of Bold Behaviors by Coat Color",
       fill = "Color") +
  theme_void()

# Visualize the relationship between coat color and timidness - Pie chart
ggplot(timid_summary_pie, aes(x = "", y = total_count, fill = color)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(title = "Proportion of Timid Behaviors by Coat Color",
       fill = "Color") +
  theme_void()

# Visualize the relationship between coat color and boldness - scatterplot 
# Assign jittered points for better distribution visualization
ggplot(bold_data, aes(x = color, y = Behavior, color = color)) +
  geom_jitter(width = 0.2, height = 0.2, size = 3) +
  labs(title = "Scatter Plot of Bold Behaviors by Coat Color",
       x = "Coat Color",
       y = "Bold Behavior") +
  theme_minimal() +
  theme(legend.position = "none")

# Visualize the relationship between coat color and timidness - scatterplot 
# Assign jittered points for better distribution visualization
ggplot(timid_data, aes(x = color, y = Behavior, color = color)) +
  geom_jitter(width = 0.2, height = 0.2, size = 3) +
  labs(title = "Scatter Plot of Timid Behaviors by Coat Color",
       x = "Coat Color",
       y = "Timid Behavior") +
  theme_minimal() +
  theme(legend.position = "none")
```

##Data visualization of bold and timid behaviors by coat color and gender
```{r}
# Add a classification column for bold and timid behaviors
activity_budget_sheet <- activity_budget_sheet %>%
  mutate(BehaviorType = case_when(
    Behavior %in% bold_behaviors ~ "Bold",
    Behavior %in% timid_behaviors ~ "Timid",
    TRUE ~ NA_character_
  ))

# Filter data for only bold and timid behaviors
filtered_data <- activity_budget_sheet %>%
  filter(!is.na(BehaviorType))

# Visualization 1: Stacked Bar Chart (Coat Color vs Bold/Timid Behaviors)
ggplot(filtered_data, aes(x = color, fill = BehaviorType)) +
  geom_bar(position = "stack") +
  facet_wrap(~ sex) +
  labs(title = "Bold and Timid Behaviors by Coat Color and Sex",
       x = "Coat Color",
       y = "Count of Behaviors",
       fill = "Behavior Type") +
  theme_minimal()

# Visualization 2: Grouped Bar Chart (Coat Color vs Behavior Type)
ggplot(filtered_data, aes(x = color, fill = BehaviorType)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ sex) +
  labs(title = "Distribution of Bold and Timid Behaviors by Coat Color and Sex",
       x = "Coat Color",
       y = "Count of Behaviors",
       fill = "Behavior Type") +
  theme_minimal()
```
##Visualizations of bold and timid behavior by coat color, sex, and age
```{r}
# Ensure Age is numeric for proper ordering in scatter plot
filtered_data <- filtered_data %>%
  mutate(age = as.numeric(age))

# Visualization 1: Faceted Scatter Plot (Age vs Behavior Type, by Sex and Coat Color)
ggplot(filtered_data, aes(x = age, y = BehaviorType, color = color)) +
  geom_jitter(width = 0.3, height = 0.2, size = 3) +
  facet_wrap(~ sex) + scale_x_continuous(breaks = seq(0, 16, 1)) + # Ensure all ages from 0 to 16 are shown
  labs(title = "Age and Behavior Type (Bold vs Timid) by Sex and Coat Color",
       x = "Age",
       y = "Behavior Type",
       color = "Coat Color") +
  theme_minimal()

```
##Count how many behaviors 'Guest in enclosure', 'Guest outside enclosure', 'Handler in enclosure' and 'Handler outside enclosure' where the response value is 'yes' are bold or timid behaviors
```{r}
# Define columns of interest
columns_of_interest <- c("Guest in enclosure", 
                         "Guest outside enclosure", 
                         "Handler in enclosure", 
                         "Handler outside enclosure")

# Filter rows where any of the columns of interest have a "yes" value
relevant_data_yes <- activity_budget_sheet %>%
  filter(if_any(all_of(columns_of_interest), ~ . == "yes")) %>%
  filter(!is.na(BehaviorType)) # Keep only rows classified as Bold or Timid

# Filter rows where any of the columns of interest have a "no" value
relevant_data_no <- activity_budget_sheet %>%
  filter(if_any(all_of(columns_of_interest), ~ . == "no")) %>%
  filter(!is.na(BehaviorType)) # Keep only rows classified as Bold or Timid

# Count rows with "yes" for each column, grouped by BehaviorType
counts_yes <- relevant_data_yes %>%
  pivot_longer(cols = all_of(columns_of_interest), 
               names_to = "Column", 
               values_to = "Value") %>%
  filter(Value == "yes") %>%
  group_by(BehaviorType, Column) %>%
  summarise(Count = n(), .groups = "drop")

# Count rows with "no" for each column, grouped by BehaviorType
counts_no <- relevant_data_no %>%
  pivot_longer(cols = all_of(columns_of_interest), 
               names_to = "Column", 
               values_to = "Value") %>%
  filter(Value == "no") %>%
  group_by(BehaviorType, Column) %>%
  summarise(Count = n(), .groups = "drop")


# View the result
print(counts_yes)
####28 observations where either the guest or handler were outside the enclosure and the wolf exhibited 'bold' behavior
####3 observations where either the guest or handler were outside the enclosure and the wolf exhibited 'timid' behavior
print(counts_no)
####majority of observations had no humans present 
```
